\textsf{Are large language models superhuman chemists?}

### 2024-10-21 src/tex/appendix.tex
add Bloom taxonomy ref
polishing

        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 919 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 987 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By measuring atomic force microscopy \\
        & & B. By measuring fourier transform infrared spectroscopy \\
        & & C. By measuring photoluminescence spectrum \\
        & & D. By measuring the absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC\-(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 128 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule

@@

\clearpage
\subsection{Model performance}
We also evaluated the model performance on the entire \chembench corpus.
\Cref{fig:barplot_all_correct_all_questions} shows the fraction of questions that were answered correctly by the models.
Note that this ranking differs from the one on the \enquote{tiny} subset

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/correlation_plot_is_number_nmr_peaks_complexity.pdf}
    \caption{\textbf{Dependence of the mean absolute error in predicting the number of NMR signals on the Böttcher complexity of the molecules.} The complexity measure proposed by \textcite{B_ttcher_2016} is an information-theoretic additive measure of compound complexity that follows chemical intuitions.
    The plot shows that for the \glspl{llm}, the predictive performance (measured as the mean absolute error in the prediction of the number of NMR signals) is not correlated with the complexity of the molecules. For inference based on reasoning, one would expect that the complexity of the molecule is a good predictor of the difficulty of the question.}
    \script{correlate_with_molecule_features.py}
    \label{fig:correlation_plot_is_number_nmr_peaks_complexity}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/correlation_plot_is_number_nmr_peaks_num_atoms.pdf}
    \caption{\textbf{Dependence of the mean absolute error in predicting the number of NMR signals on the number of atoms.} The plot shows that for the \glspl{llm}, the predictive performance (measured as the mean absolute error in the prediction of the number of NMR signals) is correlated with the number of atoms in the molecule.
    For reasoning-based inference, one would expect that the number of atoms in the molecules is not necessarily a good predictor, and certainly worse than complexity measures, of the difficulty of the question.}
    \script{correlate_with_molecule_features.py}
    \label{fig:correlation_plot_is_number_nmr_peaks_num_atoms}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/correlation_plot_is_electron_counts_num_atoms.pdf}
    \caption{\textbf{Dependence of the mean absolute error in predicting total electron counts on the number of atoms.} The plot shows that for the \glspl{llm}, the predictive performance (measured as the mean absolute error in the prediction of the total electron counts) is correlated with the number of atoms in the molecule.}
    \script{correlate_with_molecule_features.py}
    \label{fig:correlation_plot_is_electron_counts_num_atoms}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/model_size_plot.pdf}
    \caption{\textbf{Performance of models as a function of model size.} The plot shows the performance of the models as a function of the parameter count. The performance is measured as the fraction of questions answered correctly by the models. We see that the performance of the models correlates with scale for the models of the LLama3 and Llama3.1 herd of models.}
    \script{performance_vs_model_size.py}
    \label{fig:model_size_plot}
\end{figure}

@@

\clearpage
\subsection{LLM Parsing} \label{sec:llm-parsing}

@@

\begin{table}
    \centering
    \caption{\textbf{Refusal counts and parsing.} The table shows the number of refusals detected and the number of times the fallback mechanism was used for each model.}
    \variable{output/model_refusal_table.tex}
    \label{tab:refusal_counts_and_parsing}
\end{table}

@@

In the applications, human participants were presented with molecules as rendered drawings and SMILES strings. \LaTeX\xspace equations and chemical equations were rendered using MathJax.

@@

Additionally, we prompted users to provide additional information about their experience in chemistry.
While we recorded fine-grained information, e.g., their specialization, we focused on the number of years since the first university-level chemistry course.
\Cref{fig:experience_vs_correctness} shows that the experience of the human scorers was correlated with the correctness of their answers (\Cref{fig:experience_vs_correctness}, Spearman's \(\rho \approx \variable{output/spearman_experience_score.txt}\), and \(p \approx \variable{output/spearman_experience_score_p.txt}\)).

@@

Our analysis of both log probabilities and prompting confidence reveals distinct calibration behaviors across different language models.
\GPTFourO demonstrates an overconfident tendency, often assigning high probabilities even to incorrect answers.
However, when \GPTFourO displays high confidence, it accurately predicts correct answers approximately 80\% of the time. In contrast, \LlamaThreeOneEightBInstruct confidence distribution is more evenly spread, with a majority of predictions centered around 0.5. High-confidence predictions from \LlamaThreeOneEightBInstruct are less frequent compared to \GPTFourO, and unlike \GPTFourO, high confidence does not necessarily correlate with a higher chance of correct answers.

@@

\clearpage
\subsection{Leaderboard}
Our leaderboard is based on the tool chain developed for Matbench.\autocite{Dunn_2020}
Briefly, the \chembench pipeline produces standardized files in \texttt{json} format that contributors can add via pull requests to the \chembench repository.
The Markdown tables and interactive plots are automatically generated and updated on the \chembench website.

>>>>

        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 919 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 987 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By atomic force microscopy \\
        & & B. By transform infrared spectroscopy \\
        & & C. By measuring a photoluminescence spectrum \\
        & & D. By absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC\-(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 128 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule

@@

\clearpage
\subsection{Model performance}
We also evaluated the model performance on the entire \chembench corpus.
\Cref{fig:barplot_all_correct_all_questions} shows the fraction of questions that were answered correctly by the models.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/correlation_plot_is_number_nmr_peaks_complexity.pdf}
    \caption{\textbf{Dependence of the mean absolute error in predicting the number of NMR signals on the Böttcher complexity of the molecules.} The complexity measure proposed by \textcite{B_ttcher_2016} is an information-theoretic additive measure of compound complexity that follows chemical intuitions.
    The plot shows that for the \glspl{llm}, the predictive performance (measured as the mean absolute error in the prediction of the number of NMR signals) is not correlated with the complexity of the molecules. For inference based on reasoning, one would expect that the complexity of the molecule is a good predictor of the difficulty of the question.}
    \script{correlate_with_molecule_features.py}
    \label{fig:correlation_plot_is_number_nmr_peaks_complexity}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/correlation_plot_is_number_nmr_peaks_num_atoms.pdf}
    \caption{\textbf{Dependence of the mean absolute error in predicting the number of NMR signals on the number of atoms.} The plot shows that for the \glspl{llm}, the predictive performance (measured as the mean absolute error in the prediction of the number of \gls{nmr} signals).}
    \script{correlate_with_molecule_features.py}
    \label{fig:correlation_plot_is_number_nmr_peaks_num_atoms}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/correlation_plot_is_electron_counts_num_atoms.pdf}
    \caption{\textbf{Dependence of the mean absolute error in predicting total electron counts on the number of atoms.} The plot shows that for the \glspl{llm}, the predictive performance (measured as the mean absolute error in the prediction of the total electron counts) is sometimes correlated with the number of atoms in the molecule.}
    \script{correlate_with_molecule_features.py}
    \label{fig:correlation_plot_is_electron_counts_num_atoms}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/model_size_plot.pdf}
    \caption{\textbf{Performance of models as a function of model size.} The plot shows the performance of the models as a function of the parameter count. The performance is measured as the fraction of questions answered correctly by the models. We see that the performance of the models correlates with scale for the models of the LLama-3 and Llama-3.1 herd of models.}
    \script{performance_vs_model_size.py}
    \label{fig:model_size_plot}
\end{figure}

@@

\subsection{LLM Parsing} \label{sec:llm-parsing}

@@

\begin{table}
    \centering
    \caption{\textbf{Refusal counts and parsing.} The table shows the number of refusals detected and the number of times the \gls{llm} fallback parsing mechanism was used for each model.}
    \variable{output/model_refusal_table.tex}
    \label{tab:refusal_counts_and_parsing}
\end{table}

@@

In the applications, human participants were presented with molecules as rendered drawings and SMILES strings. \LaTeX\xspace equations and chemical equations were rendered using MathJax (\Cref{fig:screenshots}.

\begin{figure}
    \subfloat[A physcial chemistry question.]{
        \includegraphics[width=.45\textwidth]{figures/screenshot_a.png}
    }
    \subfloat[An organic chemistry question.]{
        \includegraphics[width=.45\textwidth]{figures/screenshot_b.png}
    }
    \caption{\textbf{Examples of how questions were shown to the human participants.}}
    \label{fig:screenshots}
\end{figure}

@@

Additionally, we prompted users to provide additional information about their experience in chemistry.
While we recorded fine-grained information, e.g., their specialization, we focused on the number of years since the first university-level chemistry course.
\Cref{fig:experience_vs_correctness} shows that the experience of the human scorers was weakly correlated with the correctness of their answers (\Cref{fig:experience_vs_correctness}, Spearman's \(\rho \approx \variable{output/spearman_experience_score.txt}\), and \(p \approx \variable{output/spearman_experience_score_p.txt}\)).

@@

Our analysis of both log probabilities and prompting confidence reveals distinct calibration behaviors across different language models.
\GPTFourO demonstrates an overconfident tendency, often assigning high probabilities even to incorrect answers.
However, when \GPTFourO displays high confidence, it accurately predicts correct answers approximately \SI{80}{\percent} of the time. In contrast, \LlamaThreeOneEightBInstruct confidence distribution is more evenly spread, with a majority of predictions centered around 0.5. High-confidence predictions from \LlamaThreeOneEightBInstruct are less frequent compared to \GPTFourO, and unlike \GPTFourO, high confidence does not necessarily correlate with a higher chance of correct answers.

@@

\clearpage
\subsection{Leaderboard}
Our leaderboard is based on the tool chain developed for Matbench.\autocite{Dunn_2020}
Briefly, the \chembench pipeline produces standardized files in \texttt{json} format that contributors can add via pull requests to the \chembench repository.
The Markdown tables and interactive plots are automatically generated and updated on the \chembench website. The leaderboard is available at \url{https://lamalab-org.github.io/chem-bench/leaderboard/}.

### 2024-10-21 src/tex/ms.tex
polishing

 These findings also shine an interesting light on the value of textbook-inspired questions.
 A subset of the questions in the \chembench are based on textbooks targeted at undergraduate students.
 On those questions, the models tend to perform better than on some of our semi-automatically constructed tasks (see \Cref{fig:performance_per_topic}).
 For instance, while the overall performance in the chemical safety topic is low, the models would pass the certification exam according to the German Chemical Prohibition Ordinance based on a subset of questions we sampled from the corresponding question bank (e.g., \variable{output/subset_scores/is_gfk_gpt-4.txt}\% correct answers for \GPTFour, \variable{output/subset_scores/is_gfk_Claude-3.5_(Sonnet).txt}\% for \ClaudeThreeFiveSonnet, and \variable{output/human_subset_scores/is_gfk.txt}\% for the human experts).
 While those findings are impacted by the subset of questions we sampled, the results still highlight that good performance on such question bank or textbook questions does not necessarily translate to good performance on other questions that require more reasoning. The findings also underline that such exams might have been a good surrogate for general performance of skills for humans, but their applicability in the face of systems that can consume vast amounts of data is up for debate.

@@

It is important to note that the model performance for some topics, however, is slightly underestimated in the current evaluation.
 This is because models provided via \glspl{api} typically have safety mechanisms that prevent them from providing answers that the provider deems unsafe.
 For instance, models might refuse to provide answers about cyanides. Statistics of the frequency of such refusals are shown in \Cref{tab:refusals}.
 To overcome this, direct access to the model weights would be required, and we strive to collaborate with the developers of frontier models to overcome this limitation in the future.
 This is facilitated with the tooling \chembench provides, thanks to which contributors can automatically add new models in an open science fashion.

@@

One interesting finding of recent research is that foundation models can judge interestingness.\autocite{zhang2024omniopenendednessmodelshuman, Argyle_2023}
If models could do so for chemical compounds, this would open opportunities for novel optimization approaches.
Such open-ended tasks, however, depend on an external observer defining what interestingness is.\autocite{hughes2024openendednessessentialartificialsuperhuman}
Here, we posed models the same question~\textcite{Choung_2023} asked chemists at a drug company: \enquote{Which of the two compounds do you prefer?} (in the context of an early virtual screening campaign setting) (see \Cref{tab:chembench_corpus_cognitive}).
While chemists showed a fair degree of interrater agreement, most of our models do not correlate with the preferences of expert chemists---even if they perform well on many other tasks in \chembench (\Cref{tab:performance_table_human_subset})
This indicates that using preference tuning for chemical settings is a promising approach to explore in future research.

@@

 \begin{figure}[!h]
     \centering
     \includegraphics[width=\textwidth]{figures/model_confidence_performance.pdf}
     \caption{\textbf{Calibration and distribution of confidence estimates.} For this analysis, we used verbalized confidence estimates from the model. We prompted the models to return a confidence score on an ordinal scale to obtain those estimates.
     The line plot shows the average fraction of correctly answered questions for each confidence level. The bar plot shows the distribution of confidence estimates. A confidence estimate would be well-calibrated if the average fraction of correctly answered questions increases with the confidence level.
     We find that the models are not well-calibrated and some provide misleading confidence estimates.
     }
     \label{fig:confidence_vs_performance}
     \script{confidence_estimate.py}
 \end{figure}

 In \Cref{fig:confidence_vs_performance}, we show that for some models, there is no significant correlation between the estimated difficulty and whether the models answered the question correctly or not.
 For applications in which humans might rely on the models to provide answers, this is a concerning observation highlighting the need for critical reasoning in the interpretation of the model's outputs.\autocite{Li_2023, miret2024llms}
 For example, for the questions about the safety profile of compounds, \GPTFour reported an average confidence of \variable{output/model_confidence_performance/gpt-4_is_pictograms_average_confidence_correct_overall.txt} (on a scale of 1--5) for the \variable{output/model_confidence_performance/gpt-4_is_pictograms_num_correct_overall.txt} questions it answered correctly and \variable{output/model_confidence_performance/gpt-4_is_pictograms_average_confidence_incorrect_overall.txt} for the \variable{output/model_confidence_performance/gpt-4_is_pictograms_num_incorrect_overall.txt} questions it answered incorrectly.
 While, on average, the verbalized confidence estimates from Claude 3 seem better calibrated (\Cref{fig:confidence_vs_performance}), they are still misleading in some cases.
 For example, for the questions about \gls{ghs} pictograms Claude 3 returns an average score of \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_correct_overall.txt} for correct answers and \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_incorrect_overall.txt} for incorrect answers.

\section{Conclusions}
On the one hand, our findings underline the impressive capabilities of \glspl{llm} in the chemical sciences: Leading models outperform domain experts in specific chemistry questions on many topics.
On the other hand, there are still striking limitations.
For very relevant topics the answers models provide are wrong.
On top of that, many models are not able to reliably estimate their own limitations.
Yet, the success of the models in our evaluations perhaps also reveals more about the limitations of the exams we use to evaluate models---and chemists---than about the models themselves.
For instance, while models perform well on many textbook questions, they struggle with questions that require some more reasoning about chemical structures (e.g., number of isomers or \gls{nmr} peaks).
Given that the models outperformed the average human in our study, we need to rethink how we teach and examine chemistry.
Critical reasoning is increasingly essential, and rote solving of problems or memorization of facts is a domain in which \glspl{llm} will continue to outperform humans.

@@

\subsection{Curation workflow}\label{sec:curation}
For our dataset, we curated questions from existing exams or exercise sheets but also programmatically created new questions.
Questions were added via Pull Requests on our GitHub repository and only merged into the corpus after passing manual review (\Cref{fig:curation_workflow}) as well as automated checks (e.g., for compliance with a standardized schema).

@@

@@

\begin{table}
    \centering
    \caption{\textbf{Overview of sources of the curated questions}. The table provides an overview of the types of sources the questions have been curated from. Detailed sources are available in the source data. Questions without source have been curated completely from scratch. Questions based on lecture notes or URL have been curated based on content presented in those resources. All questions have been rephrased, annotated, and reviewed before being added to the corpus.}
    \variable{output/sources_table.tex}
    \label{tab:sources}
\end{table}

@@

\subparagraph{Chemical preference data}

@@

To understand whether the capabilities of the leading models align with the preferences of professional chemists, we randomly selected 1000 data points from the original dataset to create a meaningful evaluation set, where molecules are represented as SMILES.
To ablate the effect of different molecular representations, we only considered questions for which we could obtain IUPAC names for both molecules present.

\subsection{Model evaluation workflow}
An overview of the pipeline is shown in \Cref{fig:process}.

@@

\subsection{Confidence estimate}
To estimate the models' confidence, we prompted them with the question (and answer options for \gls{mcq}) and the task to rate their confidence to produce the correct answer on a scale from 1 to 5.
We decided to use verbalized confidence estimates\autocite{xiong2023llms} since we found those closer to current practical use cases than other prompting strategies, which might be more suitable when implemented in systems. In addition, this approach captures semantic uncertainty which is not the same as the probability of a token given a sequence of tokens (i.e., the uncertainty one obtains from logit-based approaches). In \Cref{sec:confidence_estimates} we provide more details and comparisons with a logit-based approach.

@@

\paragraph{Study design}
Human participants were asked the questions in a custom-built web interface (\Cref{sec:human_baseline}) which rendered chemicals and equations. Questions were shown in random order and participants were not allowed to skip questions. For a subset of the questions, the participants were allowed to use external tools (excluding other \gls{llm} or asking other people) to answer the questions. Prior to answering questions, participants were asked to provide information about their education and experience in chemistry. The study was conducted in English.

\paragraph{Participants}
Users were open to reporting about their experience in chemistry.
Overall, \variable{output/num_users_with_education_info.txt} did so.
Out of those,
\variable{output/num_human_postdoc.txt} are beyond a first postdoc, \variable{output/num_human_master.txt} have a master's degree (and are currently enrolled in Ph.D.\ studies), and \variable{output/num_human_bachelor.txt} have a bachelor's degree.

\paragraph{Comparison with models}
For the analysis, we treated each human as a model. We computed the topic aggregated averages per human for analyses grouped by topic and then averaged over all humans.

@@

\section*{Acknowledgements}
This work was supported by the Carl Zeiss Foundation, a \enquote{Talent Fund} of the \enquote{Life} profile line of the Friedrich Schiller University Jena.
We also thank Stability.AI for the access to its HPC cluster. In addition, M.S-W.'s work was supported by Intel and Merck via the AWASES programme.
K.M.J.\ is part of the NFDI consortium FAIRmat funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – project 460197019.
Parts of A.M.'s work was supported as part of the \enquote{SOL-AI} project funded by the Helmholtz Foundation model initative.

>>>>

 These findings also shine an interesting light on the value of textbook-inspired questions.
 A subset of the questions in the \chembench are based on textbooks targeted at undergraduate students.
 On those questions, the models tend to perform better than on some of our semi-automatically constructed tasks (see \Cref{fig:performance_per_topic}).
 For instance, while the overall performance in the chemical safety topic is low, the models would pass the certification exam according to the German Chemical Prohibition Ordinance based on a subset of questions we sampled from the corresponding question bank (e.g., \variable{output/subset_scores/is_gfk_gpt-4.txt}\% correct answers for \GPTFour, \variable{output/subset_scores/is_gfk_Claude-3.5_(Sonnet).txt}\% for \ClaudeThreeFiveSonnet, and \variable{output/human_subset_scores/is_gfk.txt}\% for the human experts).
 While those findings are impacted by the subset of questions we sampled, the results still highlight that good performance on such question bank or textbook questions does not necessarily translate to good performance on other questions that require more reasoning or are further away from the training corpus.\autocite{mccoy2023embersautoregressionunderstandinglarge} The findings also underline that such exams might have been a good surrogate for general performance of skills for humans, but their applicability in the face of systems that can consume vast amounts of data is up for debate.

@@

It is important to note that the model performance for some topics, however, is slightly underestimated in the current evaluation.
 This is because models provided via \glspl{api} typically have safety mechanisms that prevent them from providing answers that the provider deems unsafe.
 For instance, models might refuse to provide answers about cyanides. Statistics of the frequency of such refusals are shown in \Cref{tab:refusal_counts_and_parsing}.
 To overcome this, direct access to the model weights would be required, and we strive to collaborate with the developers of frontier models to overcome this limitation in the future.
 This is facilitated with the tooling \chembench provides, thanks to which contributors can automatically add new models in an open science fashion.

@@

One interesting finding of recent research is that foundation models can judge interestingness or human preferences in some domains.\autocite{zhang2024omniopenendednessmodelshuman, Argyle_2023}
If models could do so for chemical compounds, this would open opportunities for novel optimization approaches.
Such open-ended tasks, however, depend on an external observer defining what interestingness is.\autocite{hughes2024openendednessessentialartificialsuperhuman}
Here, we posed models the same question~\textcite{Choung_2023} asked chemists at a drug company: \enquote{Which of the two compounds do you prefer?} (in the context of an early virtual screening campaign setting, see \Cref{tab:chembench_corpus_cognitive} for an example).
Despite chemists demonstrating a reasonable level of interrater agreement, our models largely fail to align with expert chemists' preferences. Their performance is often indistinguishable from random guessing, even though these same models excel in various other tasks in \chembench (\Cref{tab:performance_table_human_subset}).
This indicates that using preference tuning for chemical settings is a promising approach to explore in future research.

@@

 \begin{figure}[!h]
     \centering
     \includegraphics[width=\textwidth]{figures/model_confidence_performance.pdf}
     \caption{\textbf{Calibration and distribution of confidence estimates.} For this analysis, we used verbalized confidence estimates from the model. We prompted the models to return a confidence score on an ordinal scale to obtain those estimates.
     The line plot shows the average fraction of correctly answered questions for each confidence level. The bar plot shows the distribution of confidence estimates. A confidence estimate would be well-calibrated if the average fraction of correctly answered questions increases with the confidence level.
     We find that most models are not well-calibrated and provide misleading confidence estimates.
     }
     \label{fig:confidence_vs_performance}
     \script{confidence_estimate.py}
 \end{figure}

 In \Cref{fig:confidence_vs_performance}, we show that for some models, there is no significant correlation between the estimated difficulty and whether the models answered the question correctly or not.
 For applications in which humans might rely on the models to provide answers with trustworthy uncertainty estimates, this is a concerning observation highlighting the need for critical reasoning in the interpretation of the model's outputs.\autocite{Li_2023, miret2024llms}
 For example, for the questions about the safety profile of compounds, \GPTFour reported an average confidence of \variable{output/model_confidence_performance/gpt-4_is_pictograms_average_confidence_correct_overall.txt} (on a scale of 1--5) for the \variable{output/model_confidence_performance/gpt-4_is_pictograms_num_correct_overall.txt} questions it answered correctly and \variable{output/model_confidence_performance/gpt-4_is_pictograms_average_confidence_incorrect_overall.txt} for the \variable{output/model_confidence_performance/gpt-4_is_pictograms_num_incorrect_overall.txt} questions it answered incorrectly.
 While, on average, the verbalized confidence estimates from Claude 3 seem better calibrated (\Cref{fig:confidence_vs_performance}), they are still misleading in some cases.
 For example, for the questions about \gls{ghs} pictograms Claude 3 returns an average score of \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_correct_overall.txt} for correct answers and \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_incorrect_overall.txt} for incorrect answers.

\section{Conclusions}
On the one hand, our findings underline the impressive capabilities of \glspl{llm} in the chemical sciences: Leading models outperform domain experts in specific chemistry questions on many topics.
On the other hand, there are still striking limitations.
For very relevant topics the answers models provide are wrong.
On top of that, many models are not able to reliably estimate their own limitations.
Yet, the success of the models in our evaluations perhaps also reveals more about the limitations of the questions we use to evaluate models---and chemists---than about the models themselves.
For instance, while models perform well on many textbook questions, they struggle with questions that require some more reasoning about chemical structures (e.g., number of isomers or \gls{nmr} peaks).
Given that the models outperformed the average human in our study, we need to rethink how we teach and examine chemistry.
Critical reasoning is increasingly essential, and rote solving of problems or memorization of facts is a domain in which \glspl{llm} will continue to outperform humans (when trained on the right training corpus).

@@

\subsection{Curation workflow}\label{sec:curation}
For our dataset, we curated questions from existing exams or exercise sheets but also programmatically created new questions (\Cref{tab:sources}).
Questions were added via Pull Requests on our GitHub repository and only merged into the corpus after passing manual review (\Cref{fig:curation_workflow}) as well as automated checks (e.g., for compliance with a standardized schema).

@@

\begin{table}
    \centering
    \caption{\textbf{Overview of sources of the curated questions}. The table provides an overview of the types of sources the questions have been curated from. Detailed sources are available in the source data on GitHub. Questions without source have been curated completely from scratch. Questions based on lecture notes or URL have been curated based on content presented in those resources. All questions have been rephrased, annotated, and reviewed before being added to the corpus.}
    \variable{output/sources_table.tex}
    \label{tab:sources}
\end{table}

@@

@@

\paragraph{Chemical preference data}

@@

To understand whether the capabilities of the leading models align with the preferences of professional chemists, we randomly selected 1000 data points from the original dataset to create a meaningful evaluation set, where molecules are represented as \gls{smiles}.
To ablate the effect of different molecular representations, we only considered questions for which we could obtain \gls{iupac} names for both molecules present.

\subsection{Model evaluation workflow}
A graphical overview of the pipeline is shown in \Cref{fig:process}.

@@

\subsection{Confidence estimate}
To estimate the models' confidence, we prompted them with the question (and answer options for \gls{mcq}) and the task to rate their confidence to produce the correct answer on a scale from 1 to 5.
We decided to use verbalized confidence estimates\autocite{xiong2023llms} since we found those closer to current practical use cases than other prompting strategies, which might be more suitable when implemented in systems. In addition, this approach captures semantic uncertainty which is not the same as the probability of a token given a sequence of tokens (i.e., the uncertainty one obtains from logit-based approaches). On top of that, many propietary models do not provide access to the logits, making this approach more general.
In \Cref{sec:confidence_estimates} we provide more details and comparisons with a logit-based approach.

@@

\paragraph{Study design}
Human participants were asked the questions in a custom-built web interface (see \Cref{sec:human_baseline}) which rendered chemicals and equations. Questions were shown in random order and participants were not allowed to skip questions. For a subset of the questions, the participants were allowed to use external tools (excluding other \gls{llm} or asking other people) to answer the questions. Prior to answering questions, participants were asked to provide information about their education and experience in chemistry. The study was conducted in English.

\paragraph{Participants}
Users were open to reporting about their experience in chemistry.
Overall, \variable{output/num_users_with_education_info.txt} did so.
Out of those,
\variable{output/num_human_postdoc.txt} are beyond a first postdoc, \variable{output/num_human_master.txt} have a master's degree (and are currently enrolled in Ph.D.\ studies), and \variable{output/num_human_bachelor.txt} have a bachelor's degree. For the analysis, we excluded participants with less than two years of experience in chemistry after their first university-level course in chemistry.

\paragraph{Comparison with models}
For the analysis, we treated each human as a model. We computed the topic aggregated averages per human for analyses grouped by topic and then averaged over all humans. The performance metrics reported for models in the main text are computed on the same questions the humans answered. Metrics for the entire corpus are reported in the appendix.

@@

\section*{Acknowledgements}
This work was supported by the Carl Zeiss Foundation, a \enquote{Talent Fund} of the \enquote{Life} profile line of the Friedrich Schiller University Jena.
We also thank Stability.AI for the access to its HPC cluster. In addition, M.S-W.'s work was supported by Intel and Merck via the AWASES programme.
K.M.J.\ is part of the NFDI consortium FAIRmat funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – project 460197019.
K.M.J.\ thanks FutureHouse (a non-profit research organization supported by the generosity of Eric and Wendy Schmidt) for supporting \PaperQATwo run by access to the \gls{api}.
Parts of A.M.'s work was supported as part of the \enquote{SOL-AI} project funded by the Helmholtz Foundation model initative.
